---
title: "Introduction to R."
author: "Sergi Mayol"
date: "21746 Data Mining 2023-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this course you will learn how to turn raw data into understanding, insight, and knowledge. This workshop's goal is to present the most basic tools in R that will allow you to discover knowledge in data. We will start by talking about datasets in general, learn how to upload them and review different procedures that we can do with them. 

## The starting point: the Data

First you must import your data into R. This typically means that you take data stored in a file, database, or web application programming interface, and load it into a data frame in R. If you can’t get your data into R, you can’t do data science nor knowledge discovery on it!

Once you’ve imported your data, it is a good idea to tidy it. Tidying your data means storing it in a consistent form that matches the semantics of the dataset with the way it is stored. In brief, when your data is tidy, each column is a variable, and each row is an observation. Tidy data is important because the consistent structure lets you focus your struggle on questions about the data, not fighting to get the data into the right form for different functions.

Once you have tidy data, a common first step is to transform it. Transformation includes narrowing in on observations of interest (like all people in one city, or all data from the last year), creating new variables that are functions of existing variables (like computing speed from distance and time), and calculating a set of summary statistics (like counts or means). Together, tidying and transforming are called wrangling, because getting your data in a form that’s natural to work with often feels like a fight!

Once you have tidy data with the variables you need, we will consider its visualisation and summary. A good visualisation will show you things that you did not expect, or raise new questions about the data. A good visualisation might also hint that you’re asking the wrong question, or you need to collect different data. Visualisations can surprise you, but don’t scale particularly well because they require a human to interpret them. Summarizing data is complementary to visualisation, and gives us a general feeling of the data. 

An important part of any data analysis project is communication. It doesn’t matter how well your summaries and visualisation have led you to understand the data unless you can also communicate your results to others.

## What you need

In order to run the code provided you will need: R, RStudio, and a collection of R packages. Packages are the fundamental units of reproducible R code. They include reusable functions, the documentation that describes how to use them, and sample data.

At Aula Digital you will find resources that might help you learn R. 

### R

Go to the R-project web site http://www.r-project.org, download R for your operating system. To download R, go to CRAN, the comprehensive R archive network. CRAN is composed of a set of mirror servers distributed around the world and is used to distribute R and R packages. Don’t try and pick a mirror that’s close to you: instead use the cloud mirror, https://cloud.r-project.org, which automatically figures it out for you.

A new major version of R comes out once a year, I am currently using version R4.0.2. It’s a good idea to update regularly. Upgrading can be a bit of a hassle, especially for major versions, which require you to re-install all your packages, but putting it off only makes it worse.

### RStudio

RStudio is an integrated development environment, or IDE, for R programming. Download and install it from http://www.rstudio.com/download. RStudio is updated a couple of times a year. When a new version is available, RStudio will let you know. It’s a good idea to upgrade regularly so you can take advantage of the latest and greatest features. For this book, make sure you have at least RStudio 1.0.0.

When you start RStudio, you’ll see regions in the interface. For now, all you need to know is that you type R code in the console pane, and press enter to run it. 

Once you have R and R Studio installed, please do the following:

* Create a new, empty folder on your computer. Use only for this course's assignments. Download the Rmd file CS1-intro2R.Rmd and place it in the folder.

* Open R Studio only (you do not need to open R). Open a new script from RStudio and set the working directory to the folder you have just created.

* Open the Rmd document (CS1-intro2R.Rmd) in R Studio, File . . .Open File . . . (top left of window).

In order to run the Rmd document, you need to have the rmarkdown package installed.

### R Packages Installation

An R package is a collection of functions, data, and documentation that extends the capabilities of base R. Using packages is key to the successful use of R. Packages are installed with a simple line of code: 

* install.packages("packagename")

* Another option is to complete the following steps: 1. Click Tools → Install Packages. 2. Select Repository (CRAN). 3. Type the package name (or several package names, separated by comma). 4. Leave Install dependencies ticked as by default. 5. Click Install


On your own computer, install the rmarkdown package. You will not be able to use the functions, objects, and help files in a package until you load it. Once you have installed a package, you can load it with the library() function.


### Running R code

This CS1-intro2R.Rmd file provides you with different simple examples of how to run R code. It is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

In this first R Markdown document we will talk about datasets in general and how to upload them. When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this (insert chunk) and view the results by pressing Run Current Chunk (green arrow):
```{r trees}
(4 + 2) * 10
```

The result of the calculation is displayed.

## Math Expressions in R

R Markdown gives you the syntax to render complex mathematical formulas and derivations, and have them displayed very nicely. Inline LaTeX equations can be written in a pair of dollar signs using the LaTeX syntax, as $\pi r^2$ or $e^{i\pi}$. Math expressions of the display style can be written in a pair of double dollar signs 
$$f(k) = {n \choose k} p^{k} (1-p)^{n-k}$$

## Data Types in R    

Basic R data types that are of frequent occurrence in routine R calculations. 

* NUMERIC: decimal values are called numerics in R. It is the default computational data type 
```{r}
x <- 10.5 # assign a decimal value to x. R uses decimal points
x # print the value of x
class(x) # print the class name of x ("numeric")
```

* INTEGER: in order to create an integer variable in R, we invoke the "as.integer function"
```{r}
y <- as.integer(3)
y # print the value of y
class(y) # print the class name of y ("integer")
```

* LOGICAL: A logical value is often created via comparison between variables
```{r}
x <- 1
y <- 2 # sample values
z <- x > y # is x larger than y?
z # print the logical value
class(z) # print the class name of z ("logical")
```

Standard logical operations are "&" (and), "|" (or), and "!" (negation) can also be used.
```{r}
u <- TRUE
v <- FALSE
u & v # u AND v
u | v # u OR v
!u # negation of u
```

* CHARACTER: character object is used to represent string values in R. We convert objects into character values with the "as.character()" function:
```{r}
x <- as.character(3.14)
x # print the character string
class(x) # print the class name of x ("character")
```

Two character values can be concatenated with the "paste" function
```{r}
fname <- "Joe"
lname <- "Smith"
paste(fname, lname)
```

* VECTOR: A vector is a sequence of data elements of the same basic type. Members in a vector are also called components. 
```{r}
c(2, 3, 5) # Vector containing three numeric values 2, 3 and 5.
c(TRUE, FALSE, TRUE, FALSE, FALSE) ## Vector of logical values
c("aa", "bb", "cc", "dd", "ee") ## Vector of character strings
```

The "length" function gives the number of members in a vector
```{r}
length(c("aa", "bb", "cc", "dd", "ee"))
```

Vector arithmetics: arithmetic operations of vectors are performed member-by-member, 
```{r}
a <- c(1, 3, 5, 7)
b <- c(1, 2, 4, 8)
5 * a ## multiplies all components by 5, we would get a vector
a + b ## add a and b together, member by member
a - b ## Similarly for subtraction, multiplication and division, we get new vectors via
a * b ## memberwise operations.
a / b
```

* MATRICIES
```{r}
B <- matrix(
    c(2, 4, 3, 1, 5, 7),
    nrow = 3,
    ncol = 2
)
B ## B is a matrix that has 3 rows and 2 columns
```

Combining Matrices: The columns of two matrices having the same number of rows can be  combined into a larger matrix. 
```{r}
C <- matrix(
    c(7, 4, 2),
    nrow = 3,
    ncol = 1
)
C ## C has 3 rows
```

Then we can combine the columns of B and C with "cbind".
```{r}
D <- cbind(B, C)
```

Use "rbind" to combine rows
```{r}
D[3, 1] ## gives value of 3rd row, 1st column
```

* LIST: A list is a generic vector containing other objects.
```{r}
n <- c(2, 3, 5)
s <- c("aa", "bb", "cc", "dd", "ee")
b <- c(TRUE, FALSE, TRUE, FALSE, FALSE)
x <- list(n, s, b, 3) # x is a list containing copies of n, s, b and 3
```

* DATA FRAME: A data frame (df) is used for storing data tables. It is a list of vectors of  equal length. 
```{r}
n <- c(2, 3, 5, 7, 9)
s <- c("aa", "bb", "cc", "dd", "ee")
b <- c(TRUE, FALSE, TRUE, FALSE, FALSE)
df <- data.frame(s, b, n)
df ## df is a data frame containing three vectors n, s, b.
nrow(df) ## gives the number of rows
## "ncol" gives number of columns
```

R provides built-in data frames
```{r}
trees ## built-in data frame in R
?trees ## description of the data frame
```

The top line of the table, "header", contains the column names. Each horizontal line denotes a data row, it begins with the name/number of the row, and is followed by the actual data. 
Each data member of a row is called a cell. To retrieve data in a cell, enter its row and column coordinates in the single square bracket "[]" operator. 
The two coordinates are separated by a comma.
```{r}
trees[11, 2] ## cell value from the eleventh row, second column of trees
```

Instead of printing out the entire data frame, it is often desirable to preview it with the "head" or "tail" function 
```{r}
head(trees) ## First six rows of the data frame
tail(trees) ## Last six rows of the data frame
str(trees) ## Displays the internal structure of the R object
```

"trees" is a data frame with 31 observations of three numeric variables. Notice the position of the dollar sign. The dollar sign ($) is an operator that accesses the columns of the df.
```{r}
trees$Height ## trees is the data frame, and Height is the header of one of its columns
trees[[3]] ## To retrieve the third column vector of the data frame "trees" write
trees[["Volume"]] ## or the column name
summary(trees) ## "summary" is a generic function used to produce minimum and
## maximum values, quartiles and mean
```

## Basic calculations in R 

In order to perform calculations, data must be introduced. X and Y are two vectors, that can be introduced using the concatenation function c(). To add all elements of a vector  sum() function can be used. We can also give it a name so that we can refer to it later on.
```{r}
Y <- c(8, 4, 4, 6, 13)
X <- c(3, 4, 2, -2, 8)
sumY <- sum(Y)
sumX <- sum(X)
```

There is no output, but we can call it using the name we assigned to it.
```{r}
sum(X)
sum(Y)
```

We can easily calculate the mean and variance of the values in X and in Y:
$$\overline{x} = \frac{\sum_{i=1}^n x_i}{n}$$
$$s_x^2 = \frac{\sum_{i=1}^n (x_i-\overline{x})^2}{n-1}$$

```{r}
meanofX <- sumX / length(X) ## length(X) gives us the number of values in X
meanofY <- sumY / length(Y)
varianceofX <- sum((X - meanofX)^2) / (length(X) - 1)
varianceofY <- sum((Y - meanofY)^2) / (length(Y) - 1)
```


## Generating the html document

Once all your chunks run correctly, you can **knit** the Rmd file to generate the html  document. You must have the "rmarkdown" package installed. 




