---
title: "Linear Regression"
author: "Margaret Miró"
date: "21746 2023-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## The Data: trees

In this example we will be using the trees dataset available in R.
```{r trees}
summary(trees)
```

You can also embed plots:
```{r volume and height, echo=FALSE}
plot(trees$Volume ~ trees$Height,
  main = "Timber volume vs. Tree height",
  xlab = "height (ft)", ylab = "volume (sq ft)"
)
```

## Linear Regression with lm().

The regression parameters can be calculated using the "lm()" (linear model) function that estimates the slope and y-intercept of the fitted line.

**Research question 1**: How does the tree height affect its volume? Study the linear relationship.
```{r}
trees.lm <- lm(trees$Volume ~ trees$Height)
trees.lm
```

### Function for Predictions and Errors. 
Let's check if the regression line passes through the point of means $(\overline{x},\overline{y})$, that is, the predicted value for $\overline{x}$, $\beta_0 + \beta_1 \cdot \overline{x}$ and $\overline{y}$ are equal. We start by creating a function "Yhat" (predicted value).

```{r}
Yhat <- function(x, beta0, beta1) {
  beta0 + beta1 * x
}
Yhat(mean(trees$Height), trees.lm$coefficient[1], trees.lm$coefficient[2])
mean(trees$Volume)
```

We have obtained the same results. This, the regression line passes through $(\overline{x},\overline{y})$.

What can you say about the mean of the errors? Calculate it.
```{r}
mean(trees.lm$residuals)
```

### Determination coefficient.

We will find the determination coefficient, adjusted r-squared using the "lm" output.
```{r}
### Using lm" output
summary(trees.lm)$r.squared
```

## Volume as function of Girth

**Research question 2**: How does the tree girth affect its volume? Study the linear relationship. 

The results for the linear model Volume as a function of Height are not very good. Let's study Volume as a function of Girth.
```{r}
trees.lm2 <- lm(trees$Volume ~ trees$Girth)
summary(trees.lm2)
```


## Multiple Linear Regression: Volume as a function of Height and Girth

**Research question 3**: How does the tree girth and tree height affect its volume? Study the multilinearlinear relationship. 

```{r}
trees.mlm <- lm(trees$Volume ~ trees$Girth + trees$Height)
trees.mlm
```

```{r}
summary(trees.mlm)
```