---
title: "Assignment 1"
author: "Sergi Mayol"
date: "14/09/23"
output: html_document
---

Complete the following problem without using the R built-in functions.
Consider the ChickWeight dataset built into R. To see a description of this dataset, type ?ChickWeight.
a) Calculate the mean and variance of variables weight and Time.
b) Calculate the covariance between weight and Time. Interpret the result.

## Loading dataset

```{r}
dataset <- ChickWeight
```


## a) Calculate the mean and variance of variables weight and Time.

```{r}
weight <- dataset$weight
time <- dataset$Time
```

- Mean:

```{r}
my_mean <- function(data) {
  s <- 0
  cont <- 0
  for (d in data) {
    s <- s + d
    cont <- cont + 1
  }
  return(s / cont)
}
```

```{r}
my_mean(weight)
my_mean(time)
```

- Variance

```{r}
variance <- function(data) {
  mean <- my_mean(data)
  cont <- 0
  s <- 0
  for (d in data) {
    s <- s + (d - mean)^2
    cont <- cont + 1
  }
  return(s / (cont - 1))
}
```

```{r}
variance(weight)
variance(time)
```

## b) Calculate the covariance between weight and Time. Interpret the result.

```{r}
covariance <- function(data_1, data_2) {
  mean_1 <- my_mean(data_1)
  mean_2 <- my_mean(data_2)
  cont <- 0
  s <- 0
  for (d in data_1) {
    s <- s + (data_1[cont + 1] - mean_1) * (data_2[cont + 1] - mean_2)
    cont <- cont + 1
  }
  return(s / (cont - 1))
}
```

```{r}
covariance(weight, time)
```
